---
title: "Assignment 1"
author: "Kayla Choi - MA615 2021"
date: "9/20/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#include the necessary library
library(tidyverse)
```

```{r}
#source the R script that holds our functions 
source(file="hand_functions.R", echo = TRUE)
```

```{r}
# call built-in data mtcars.
data(mtcars)
```

```{r}
# Select only car models where mpg<20
mtcars_mpg2 <- mtcars[mtcars$mpg < 20,]

# Reduce the variables to mpg, cyl, disp, hp, gears
mtcars_mpg2 <- mtcars_mpg2[, c(1,2,3,4,10)]

#look at the data frame and its summary
head(mtcars_mpg2)
summary(mtcars_mpg2)
```
  
Looking at the summary of mtcars_mpg2 data frame, it is evident that the descriptive statistics for cyl (number of cylinders) and gear are not meaningful. This is because they are factor variables and not continuous.    
  
```{r}
# Now use the function from hand_functions.R and store in a variable
sp_out <- sum_special(mtcars_mpg2)
sp_out
```
   
This function outputs a list of summary statistics, including mean, variance, covariance, and correlation. It is interesting to note that there is a high negative correlation between disp and mpg. Let's explore that in the next ggplot.    
   
   
```{r}
#explore visualization of disp vs mpg
ggplot(mtcars_mpg2) +
  aes(x = disp, y = mpg) + 
  geom_point(shape = "bullet", size = 4L, colour = "#B22222") +
  geom_smooth(span = 0.5) +
  theme_bw()
```
  
From this plot, we can see that as displacement (cu.in.) increases, mpg generally decreases. However, I notice two points that do not follow this general trend. I want to look at which ones these are.    
  
```{r}
#look at the two unusual points
disp_over_350 <- mtcars_mpg2[(mtcars_mpg2$disp) >= 350, ]
unusual_mpg <- disp_over_350[(disp_over_350$mpg) >= 17, ]
unusual_mpg
```
  
They are the Hornet Sportabout and the Pontiac Firebird.   
  
  
Next, let's look at cylinders vs mpg. Because cylinder is a factor, as I mentioned above when we looked at the summary statistics, we need to make sure we convert it via as.factor().    
```{r}
#explore visualization of cyl vs mpg 
ggplot(mtcars_mpg2, aes(x=as.factor(cyl), y=mpg)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) + 
    stat_summary(fun=mean, geom="point", shape=20, size=4, color = "blue") +
    xlab("cyl") + 
    theme_bw()
```
  
For cars with 8 cylinders, there is a larger spread in the mpg and the median is lower. I have added the mean as well in the visualization. There are also two outliers with 8 cylinder cars. I want to look at which ones these are.    
  
```{r}
cyl_8 <- mtcars_mpg2[(mtcars_mpg2$cyl) ==8 , ]
outlier_cyl_large <- cyl_8[(cyl_8$mpg) >= 18.75, ]
outlier_cyl_small <- cyl_8[(cyl_8$mpg) <= 12.5, ]
outlier_cyl_large
outlier_cyl_small
```
  
The Pontiac Firebird has a high mpg compared to other 8-cylinder models.   
I only saw one distinct point that was a low outlier, but it turns out that there are actually two models that have the same mpg: Cadillac Fleetwood and Lincoln Continental.

  
  
We also saw from the sum_special function that there is a high positive correlation between cylinder and displacement. Let's explore that.    
```{r}
#explore visualization of cyl vs mpg 
ggplot(mtcars_mpg2, aes(x=as.factor(cyl), y=disp)) + 
    geom_boxplot(fill="slateblue", alpha=0.2) + 
    stat_summary(fun=mean, geom="point", shape=20, size=4, color = "blue") +
    xlab("cyl") + 
    theme_bw()
```
  
For cars with 6 cylinders, the displacement is much lower, has a smaller spread, and is left-skewed. The 8 cylinder displacement has a significantly higher median displacement and a larger spread.     
  